---
title: "Introduction to brainNetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brainNetR-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`brainNetR` is an open-source R package for reproducible analysis of brain connectivity networks.

The package provides a simple workflow for transforming brain connectivity matrices into graph representations, computing core network neuroscience metrics, and visualizing the resulting network.

This vignette presents a basic workflow using the current development version of the package.

## 1. Brain connectivity matrices

In network neuroscience, brain connectivity is commonly represented as a square matrix.

Each row and column corresponds to a brain region or region of interest (ROI), while each matrix entry represents the strength of connectivity between two regions.

For example:

```{r}
mat <- matrix(
  c(
    1.0, 0.7, 0.2,
    0.7, 1.0, 0.5,
    0.2, 0.5, 1.0
  ),
  nrow = 3,
  byrow = TRUE
)

mat
```

In this example, the matrix represents three brain regions.

## 2. Converting a connectivity matrix into a graph

The `connectivity_to_graph()` function transforms a connectivity matrix into an `igraph` object.

A connectivity threshold can be specified to retain only connections above a given magnitude.

```{r}
library(brainNetR)

graph <- connectivity_to_graph(
  mat,
  threshold = 0.3
)

graph
```

Values below the selected threshold are removed before the graph is created.

The diagonal of the matrix is also removed because self-connections are not considered in the current implementation.

## 3. Computing brain network metrics

Once the graph has been created, basic network-level metrics can be calculated with `brain_network_metrics()`.

```{r}
metrics <- brain_network_metrics(graph)

metrics
```

The current version returns:

- number of nodes;
- number of edges;
- mean degree;
- clustering coefficient;
- global efficiency;
- modularity.

These measures provide a basic description of the topology of the brain network.

## 4. Visualizing the network

The network can be visualized using `plot_brain_network()`.

```{r}
plot_brain_network(graph)
```

The current visualization is intentionally simple and is designed primarily for exploratory analysis.

## 5. Complete workflow

The basic workflow implemented by `brainNetR` can be summarized as:

```text
Brain connectivity matrix
        |
        v
connectivity_to_graph()
        |
        v
Brain network
        |
        +-----------------------+
        |                       |
        v                       v
brain_network_metrics()   plot_brain_network()
        |                       |
        v                       v
Network metrics           Network visualization
```

This allows a user to move from a connectivity matrix to a graph-based representation using a small number of commands.

## 6. Scientific scope

The current version of `brainNetR` focuses specifically on the analysis of connectivity matrices that have already been estimated.

The package does not currently perform:

- fMRI preprocessing;
- extraction of ROI time series;
- estimation of functional connectivity from raw neuroimaging data;
- machine learning classification;
- causal connectivity analysis.

These steps remain outside the scope of the initial version.

The package is intended to provide a lightweight and reproducible layer between an already-computed connectivity matrix and graph-based network analysis.

## 7. Intended users

`brainNetR` may be useful for:

- researchers in computational neuroscience;
- researchers working with functional connectivity;
- students learning network neuroscience;
- users working with fMRI-derived connectivity matrices;
- researchers preparing graph-based features for statistical or machine learning analyses.

## 8. Development status

`brainNetR` is currently under active development.

The first version intentionally focuses on a small number of functions in order to maintain a clear and reproducible workflow.

Future development may include additional metrics, validation tools, visualization options, thresholding strategies, and support for more complex brain network analyses. 
